---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(warn=-1)
```

# Selective and robust external-control (srEC) integrative estimation

<!-- badges: start -->
<!-- badges: end -->

The srEC package is to develop a data-adaptive borrowing framework to incorporate external-controls (EC) with the trial data, which could potentially facilitate the drug development. By adopting the subject-level bias framework, the comparability of each EC subject is assessed via a penalized estimation. The final integrative estimator will only incorporate the ECs with the estimated bias being zero (i.e., a comparable subset).

## Installation with `devtools`:

```{r, eval = FALSE}
devtools::install_github("Gaochenyin/SelectiveIntegrative")
```

## Example 1: data-adaptive borrowing for continuous outcomes

```{r include=FALSE}
library(SelectiveIntegrative)
set.seed(2333)
# generate data for the whole population
n_c.E <- 3000; n_e.E <- 1000
N <- n_c.E + n_e.E
X1 <- rnorm(N); X2 <- rnorm(N)
# omega: guage the unmeasured confounding for EC
omega <- 3
# generate the randomized trial population 
## generate the selection indicator for RT with expected sample size n_c.E
alpha0.opt <- uniroot(function(alpha0){
  mean(exp(alpha0 -2 * X1 - 2 * X2)/
  (1 + exp(alpha0 -2 * X1 - 2 * X2))) * N - n_c.E
}, interval = c(-50, 10))$root
eS <- exp(alpha0.opt -2 * X1 - 2 * X2)/
  (1 + exp(alpha0.opt -2 * X1 - 2 * X2))
delta <- rbinom(N, size = 1, prob = eS)
X.rt <- cbind(1, X1, X2)[delta == 1, ]
(n_c <- nrow(X.rt))
#> [1] 2986
## generate the treatment assignment with marginal probability P.A
P.A <- 2/3
eta0.opt <- uniroot(function(eta0){
  mean(exp(eta0 - X.rt%*%c(1, 1, 1))/
  (1 + exp(eta0 - X.rt%*%c(1, 1, 1)))) - P.A
}, interval = c(-50, 10))$root
eA <- exp(eta0.opt - X.rt%*%c(1, 1, 1))/
  (1 + exp(eta0.opt - X.rt%*%c(1, 1, 1)))
A.rt <- rbinom(n_c, size = 1, prob = eA)
## generate the observed outcomes for RT
Y.rt <- as.vector(1 +  X.rt%*%c(1, 1, 1) + A.rt * X.rt%*%c(0, .3, .3) + rnorm(n_c) + 
                    omega * rnorm(n_c)) # maintain a similar variation as the EC


# generate the external control population
X.ec <- cbind(1, X1, X2)[delta == 0, ]
(n_h <- nrow(X.ec))
#> [1] 1014
A.ec <- 0
## generate the observed outcomes for EC (possibly confounded)
Y.ec <- as.vector(1 +  X.ec%*%c(1, 1, 1) + omega * rnorm(n_h, mean = 1) + rnorm(n_h))
```

```{r echo=TRUE}
# omega: guage the unmeasured confounding for EC
omega <- 3
# RT data and EC data
data_rt <- list(X = X.rt, A = A.rt, Y = Y.rt)
data_ec <- list(X = X.ec, A = A.ec, Y = Y.ec)
head(data_rt); head(data_ec)
```

Now, we have generated the RT dataset `data_rt` and the EC dataset `data_ec`. We are
ready to implement our selective integrative estimation by calling `srEC()`.
```{r  echo=TRUE, results='hide'}
out <- srEC(data_rt = data_rt,
     data_ec = list(data_ec),
     method = 'glm')
```
```{r echo=TRUE}
# AIPW
print(paste('AIPW: ', round(out$est$AIPW, 3), 
      ', S.E.: ', round(out$sd$AIPW, 3)))
# ACW
print(paste('ACW: ', round(out$est$ACW, 3), 
      ', S.E.: ', round(out$sd$ACW, 3)))
# selective integrative estimation
print(paste('Our: ', round(out$est$ACW.final, 3), 
      ', S.E.: ', round(out$sd$ACW.final, 3)))
```


## Example 2: data-adaptive borrowing for survival outcomes

```{r include=FALSE}
library(SelectiveIntegrative)
library(survival)
set.seed(2333)
# generate data for the whole population
N.R1 <- 500; N.R0 <- 200; N.E <- 500
N <- N.R1 + N.R0 + N.E
# generate covaraiates
p <- 3
X <- replicate(p, rnorm(N))
# omega: guage the unmeasured confounding for EC
omega <- 0.5
# generate unmeasured confounder
U <- rnorm(N)
# generate data origin indicator
alphaR.opt <- uniroot(f = function(alpha0){
  mean(exp(alpha0 + X%*%rep(1, p) + omega * U)/{1+exp(alpha0 + X%*%rep(1, p) + omega * U)} * N) - 
    N.R1 - N.R0
}, c(-10, 10))$root

prob.R <- exp(alphaR.opt + X%*%rep(1, p) + omega * U)/
  {1+exp(alphaR.opt + X%*%rep(1, p) + omega * U)}

R <- rbinom(N, size = 1, prob = prob.R)
X.R <- X[R==1, ]; X.E <- X[R==0, ]
U.R <- U[R == 1]; U.E <- U[R == 0]
N.R <- sum(R==1)

## RCT
alphaA.opt <- uniroot(f = function(alpha0){
  mean(exp(alpha0 + X.R%*%rep(1, p))/{1+exp(alpha0 + X.R%*%rep(1, p))} * (N.R1 + N.R0)) -
    N.R1
}, c(-10, 10))$root

prob.A <- exp(alphaA.opt + X.R%*%rep(1, p))/
  {1+exp(alphaA.opt + X.R%*%rep(1, p))}

A <- rbinom(N.R, size = 1, prob = prob.A)

# update the true NR1, NR0, NE
N.R1 <- sum(A==1); N.R0 <- sum(A==0)
N.E <- sum(R==0)
# treatment effect
eta_A <- -0.5
# model for event time
f.X <- function(A, X.R, b.f, R,
                beta.X = rep(-.2, p),
                omega = 0,
                t_validity = TRUE,
                delta_T = 0){
  N <- nrow(X.R)
  # generate unmeasured confounder
  if(R){U <- rnorm(N)
  }else{U <- rnorm(N) + 1}
  # generate time incurrency bias
  bias <- delta_T * sample(c(0,5), size = N,
                           prob = c(0.5, 0.5),
                           replace = TRUE) * (1-R)
  if(R|t_validity){
    c(exp(eta_A * A +  X.R%*%beta.X + omega * U + bias))
  }else{
    1/c(exp(eta_A * A +  X.R%*%beta.X + omega * U + bias))^b.f
  }
}
## survival function
surv.fct <- function(t, A, X.R,
                     b.f, g.f, R, 
                     beta.X = rep(-.2, p),
                     omega = 0,
                     t_validity = TRUE, 
                     delta_T = 0){
  exp(-f.X(A, X.R, b.f, R, beta.X,
           omega, t_validity, delta_T) * t^g.f/b.f^g.f)
}

## hazard function
hazard.fct <- function(t, A, X.R, 
                       b.f, g.f, R, 
                       beta.X = rep(-.2, p),
                       omega = 0,
                       t_validity = TRUE, 
                       delta_T = 0){
  g.f/b.f^g.f * t^{g.f-1} * f.X(A, X.R, b.f, R, beta.X,
                                omega, t_validity, delta_T)
  # {-b^g * log(1-u)/{exp(0.5 * X1 + 0.1 * X2 - A)}}^(1/g)
}
surv.fct.inv <- function(u, A, X.R, 
                         b.f, g.f, R, 
                         beta.X = rep(-.2, p),
                         omega = 0,
                         t_validity = TRUE, 
                         delta_T = 0){
  b.f * {-log(u)/f.X(A, X.R, b.f, R, beta.X,
                     omega, t_validity, delta_T)}^{1/g.f}
}


# model for censor time
fC.X <- function(X.R){
      c(exp(X.R%*%rep(0.1, p)))
    }
surv.fct.c <- function(t, X.R, b.c, g.c){
  exp(-fC.X(X.R) * t^g.c/b.c^g.c)
}

surv.fct.c.inv <- function(u, X.R, b.c, g.c){
  b.c * {-log(u)/fC.X(X.R)}^{1/g.c}
}


b.f <<- 1
# generate the outcome T by the cumulative density function of the Weibull distribution
event.timeR <- surv.fct.inv(runif(N.R), A = A, X.R = X.R, 
                            b.f = b.f, g.f = 1,
                            R = 1, 
                            beta.X = rep(-.2, p),
                            omega = omega)
# compute the theoretical censoring probability by numerically integral
b.c <<- 1
cens.timeR <- surv.fct.c.inv(runif(N.R), X.R = X.R, 
                             b.c = b.c, g.c = 1)
## EC
event.timeE <- surv.fct.inv(runif(N.E), 
                            A = 0, X.R = X.E,
                            R = 0, 
                            beta.X = rep(-.2, p),
                            omega = omega,
                            b.f = 1, g.f = 1)
cens.timeE <- surv.fct.c.inv(runif(N.E), X.R = X.E, 
                             b.c = b.c, g.c = 1)
```

```{r echo=TRUE}
# omega: guage the unmeasured confounding for EC
omega <- 0.5
# RT data and EC data
data_rt <- data.frame(Y = pmin(event.timeR, cens.timeR),
                      D = as.numeric(event.timeR <= cens.timeR), 
                      X.R, A = A)
data_ec <- data.frame(Y = pmin(event.timeE, cens.timeE), 
                    D = as.numeric(event.timeE <= cens.timeE),
                    X.E)
head(data_rt); head(data_ec)
```

Now, we have generated the RT dataset `data_rt` and the EC dataset `data_ec` for the survival outcomes. We are ready to implement our selective integrative estimation by calling `srEC_Surv()`.
```{r  echo=TRUE, results='hide'}
out_Surv <- srEC_Surv(data_rt = data_rt,
                 data_ec = data_ec, 
                 k_grid = 1:4, 
                 X.pred.R = c('X1', 'X2', 'X3'),
                 X.pred.A = c('X1', 'X2', 'X3'),
                 X.pred.c = c('X1', 'X2', 'X3'),
                 X.pred.f = c('A', 'X1', 'X2', 'X3'))
```
```{r echo=TRUE}
# AIPW
out_Surv$aipcw.RCT
# ACW
out_Surv$aipcw
# selective integrative estimation
out_Surv$aipcw_lasso
```

## Referenced Papers: 

1. C. Gao, S. Yang\*, M. Shan, W. Ye, I. Lipkovich, D. Faries. Improving randomized controlled trial analysis with data-adaptive borrowing. *Biometrika* (2025).

2. C. Gao, S. Yang\*, M. Shan, W. Ye, I. Lipkovich, D. Faries. Doubly protected estimation for survival outcomes utilizing external controls for randomized clinical trials. *International Conference on Machine Learning (ICML)* (2025).
